Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-02-12 jso.qmd
edits in sec 6

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For simpler problems (6 dimensions), small
parameter values (20 jellyfish and a maximum of 50 tries) can already
achieve a high success rate. However, larger parameter values (i.e. 100
jellyfish and a maximum of 100 tries) are necessary for
higher-dimensional problems (8, 10, and 12 dimensions). Increasing both
parameters enhances the performance of JSO, but it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, running time, number of jellyfish and maximum
number of tries, as predictors. Long runtime is a binary variable, where
greater than 20 seconds is labelled 1. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies. As expected, the JSO success rate is higher with higher squintability and/or more jellyfish, and lower when dimension is higher (higher-dimensional optimization is more difficult). Interestingly, the JSO is unaffected by the smoothness of the index function. This is consistent with the way random search algorithms jump from value to value without taking local regularity into account, as opposed to gradient-based optimizers for example. Allowing
for more tries also does not affect success rate significantly. An increase in
squintability of 10\% increases the success rate by
`r round(exp(broom::tidy(mod1)$estimate[3]) * 10, 0)`%. As dimension
increases by one the success rate halves. Increasing the number of
jellies by 10, increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 0)`%.

```{r}
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, running time, number of jellyfish and maximum number of tries. Interestingly, squintability and dimension strongly affect jellyfish optimisation success. The number of jellies marginally affects success, but index smoothness, running longer and increasing the number of tries do not."
#| fig-pos: t
library(gtsummary)
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate = TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    long_time = "Long runtime", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of tries")) |>
  bold_p(t = 0.1)
```

>>>>

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For problems with relatively lower dimensional search spaces (4 and 6 dimensions), small parameter values (20 jellyfish and a maximum of 50 tries) are sufficient for space exploration and thus can already achieve a high success rate. Higher parameter values (i.e. 100 jellyfish and a maximum of 100 tries) enhances exploration of the search space, which is particularly beneficial in higher-dimensional problems (8, 10, and 12 dimensions) where a more complex search is necessary. However, in lower-dimensional problems, the benefit may be less pronounced, as the space is alredy relatively easy to navigate. While increasing both parameters enhances the performance of JSO, it also extends the computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

@tbl-mod-output presents the results from fitting a logistic regression
model using the proportion of success as the response and smoothness,
squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies. As expected, the JSO success rate is higher with higher squintability and/or more jellyfish, and lower when dimension is higher (higher-dimensional optimization is more difficult). Interestingly, the JSO is unaffected by the smoothness of the index function. This is consistent with the way random search algorithms jump from value to value without taking local regularity into account, as opposed to gradient-based optimizers for example. Allowing
for more tries also does not affect success rate significantly. A unit increase in squintability increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[3])-1), 0)`%. As dimension
increases by one the success rate almost halves. Increasing the number of
jellies by 10, increases the success rate by
`r round(100 * (exp(broom::tidy(mod1)$estimate[5])-1), 0)`%.

```{r}
#| label: tbl-mod-output
#| tbl-cap: "Jellyfish success rate relative to index properties and jellyfish hyper-parameters. This is the summary from a logistic regression fit to smoothness, squintability, dimension, number of jellyfish and maximum number of tries. Interestingly, squintability and dimension strongly affect jellyfish optimisation success. The number of jellies marginally affects success, but index smoothness, and increasing the number of tries do not."
#| fig-pos: t
library(gtsummary)
set_gtsummary_theme(theme_gtsummary_journal("jama"))
mod1 |> tbl_regression(exponentiate = TRUE, 
                       pvalue_fun = label_style_pvalue(digits = 3),
                       label = list(smoothness = "Smoothness", 
                                    squintability = "Squintability",
                                    d = "Dimension", 
                                    #long_time = "Long runtime", 
                                    n_jellies = "Number of jellyfish", 
                                    max_tries = "Maximum number of tries")) |>
  bold_p(t = 0.1)
```

### 2025-02-12 jso.tex
edits in sec 6

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, opacityback=0, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, toprule=.15mm, titlerule=0mm, toptitle=1mm, bottomtitle=1mm, arc=.35mm, leftrule=.75mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colback=white]

@@

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For simpler
problems (6 dimensions), small parameter values (20 jellyfish and a
maximum of 50 tries) can already achieve a high success rate. However,
larger parameter values (i.e.~100 jellyfish and a maximum of 100 tries)
are necessary for higher-dimensional problems (8, 10, and 12
dimensions). Increasing both parameters enhances the performance of JSO,
but it also extends the computational time required for the
optimisation, which can be computationally intensive when evaluating the
index function (such as scagnostic indexes) multiple times across
numerous iterations.

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, running time, number of jellyfish
and maximum number of tries, as predictors. Long runtime is a binary
variable, where greater than 20 seconds is labelled 1. The fit suggests
that the JSO success rate is only affected by squintability, dimension
and number of jellies. As expected, the JSO success rate is higher with
higher squintability and/or more jellyfish, and lower when dimension is
higher (higher-dimensional optimization is more difficult).
Interestingly, the JSO is unaffected by the smoothness of the index
function. This is consistent with the way random search algorithms jump
from value to value without taking local regularity into account, as
opposed to gradient-based optimizers for example. Allowing for more
tries also does not affect success rate significantly. An increase in
squintability of 10\% increases the success rate by 12\%. As dimension
increases by one the success rate halves. Increasing the number of
jellies by 10, increases the success rate by 6\%.

@@

\caption{\label{tbl-mod-output}Jellyfish success rate relative to index
properties and jellyfish hyper-parameters. This is the summary from a
logistic regression fit to smoothness, squintability, dimension, running
time, number of jellyfish and maximum number of tries. Interestingly,
squintability and dimension strongly affect jellyfish optimisation
success. The number of jellies marginally affects success, but index
smoothness, running longer and increasing the number of tries do not.}

>>>>

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, rightrule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, leftrule=.75mm, toptitle=1mm, colback=white, titlerule=0mm, breakable, opacityback=0, opacitybacktitle=0.6, bottomtitle=1mm, toprule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, coltitle=black]

@@

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For problems
with relatively lower dimensional search spaces (4 and 6 dimensions),
small parameter values (20 jellyfish and a maximum of 50 tries) are
sufficient for space exploration and thus can already achieve a high
success rate. Higher parameter values (i.e.~100 jellyfish and a maximum
of 100 tries) enhances exploration of the search space, which is
particularly beneficial in higher-dimensional problems (8, 10, and 12
dimensions) where a more complex search is necessary. However, in
lower-dimensional problems, the benefit may be less pronounced, as the
space is alredy relatively easy to navigate. While increasing both
parameters enhances the performance of JSO, it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

@@

Table~\ref{tbl-mod-output} presents the results from fitting a logistic
regression model using the proportion of success as the response and
smoothness, squintability, dimension, number of jellyfish and maximum
number of tries, as predictors. The fit suggests that the JSO success
rate is only affected by squintability, dimension and number of jellies.
As expected, the JSO success rate is higher with higher squintability
and/or more jellyfish, and lower when dimension is higher
(higher-dimensional optimization is more difficult). Interestingly, the
JSO is unaffected by the smoothness of the index function. This is
consistent with the way random search algorithms jump from value to
value without taking local regularity into account, as opposed to
gradient-based optimizers for example. Allowing for more tries also does
not affect success rate significantly. A unit increase in squintability
increases the success rate by 24\%. As dimension increases by one the
success rate almost halves. Increasing the number of jellies by 10,
increases the success rate by 32\%.

@@

\caption{\label{tbl-mod-output}Jellyfish success rate relative to index
properties and jellyfish hyper-parameters. This is the summary from a
logistic regression fit to smoothness, squintability, dimension, number
of jellyfish and maximum number of tries. Interestingly, squintability
and dimension strongly affect jellyfish optimisation success. The number
of jellies marginally affects success, but index smoothness, and
increasing the number of tries do not.}
